Morus: Biases in the stream

### 2018-05-09 morusAC_01_Introduction.tex
propose introduction (poorly written)

Authenticated encryption is ... \\

\noindent
CAESAR \cite{CAESAR} ...\\

@@

\subsubsection*{Paper Outline.}
This paper is organized as follows......

>>>>

Block ciphers provide confidentiality through encryption. Hash functions are the building blocks of integrity and authenticity. These two operations combine in a single scheme class named Authenticated Encryption. From a message; IV and key, it produces a ciphertext and a final authentication tag.

\noindent
The CAESAR\cite{CAESAR} Competition aims to provide with a portfolio of authenticated encryption schemes. The selection is made by comparing the resistances to a wide range of attacks. Morus is one of the last 3 finalist.

@@

\subsubsection*{Paper Outline.}
This paper is organized as follows, we first provide a brief introduction of how Morus operates\ref{sec/Preliminaries}. By showing existance of a certain class of rotational symmetries, we provide with a reduction of Morus namely MiniMorus\ref{sec/introminimorus}. We analysed this simplified scheme in \ref{sec/minitrails}, building a ciphertext-only trail with a weight of 16.
We then extend our result to the full scheme, showing a bias in the keystream over 5 rounds. In \ref{sec/IniFin} we provide new attacks on reduced version of the initialization and the finalization.

### 2018-05-09 morusAC_02_Preliminaries.tex
split text into sections
propose introduction (poorly written)

\section{Preliminaries}
\label{Sec/Preliminaries}

\subsection{Specification of MORUS}
\label{subsec/Spec}
MORUS \cite{MORUS} is a family of authenticated ciphers with two internal state sizes: 640 and 1280 bits. Each member is called \cipher{MORUS640} and \cipher{MORUS1280}, respectively. In the recommended parameters, \cipher{MORUS640} supports 128-bit keys and \cipher{MORUS1280} supports both 128-bit and 256-bit keys. The tag size is 128 bits or shorter. The designers strongly recommend using a 128-bit tag.

The state of MORUS is divided into five registers. Let $q$ be the register size, where $q = 128$ for \cipher{MORUS640} and $q=256$ for \cipher{MORUS1280}. Let $S_i, i \in\{0,1,2,3,4\}$ be a register. Then, the state is represented as $S_0\|S_1\|S_2\|S_3\|S_4$. The state is first initialized by a key and nonce, and then updated by associated data blocks and message blocks step by step. Hence, we use a superscript to denote the step number of the state as $S^t_0\|S^t_1\|S^t_2\|S^t_3\|S^t_4$, where $t$ is $-16$ before the initialization and $t=0$ after the initialization.

@@

The initialization of \cipher{MORUS1280} is slightly different from that of \cipher{MORUS640} due to the different register size and the two possible key size:
\begin{itemize}
  \item[-] When the key size is 128 bits, $K = K_{128} \|K_{128}$
  \item[-] When the key size is 256 bits, $K = K_{256}$
\end{itemize}
The state is initialized as
\begin{align*}
S^{-16}_0 &= \textrm{IV}_{128} \| 0^{128}, &
S^{-16}_1 &= K, &
S^{-16}_2 = 1^{256}, \\
S^{-16}_3 &= 0^{256}, &
S^{-16}_4 &= const_1\|const_0.
\end{align*}
After updating the state 16 times, the state is updated by $S^0_1 \leftarrow S^0_1 \oplus K$.

>>>>

\section{Preliminaries}
\label{sec/Preliminaries}

MORUS \cite{MORUS} is a family of authenticated ciphers with two internal state sizes: 640 and 1280 bits. Each member is called \cipher{MORUS640} and \cipher{MORUS1280}, respectively. In the recommended parameters, \cipher{MORUS640} supports 128-bit keys and \cipher{MORUS1280} supports both 128-bit and 256-bit keys. The tag size is 128 bits or shorter. The designers strongly recommend using a 128-bit tag.

\subsection{Specification of MORUS}
\label{subsec/Spec}
The state of MORUS is divided into five registers. Let $q$ be the register size, where $q = 128$ for \cipher{MORUS640} and $q=256$ for \cipher{MORUS1280}. Let $S_i, i \in\{0,1,2,3,4\}$ be a register. Then, the state is represented as $S_0\|S_1\|S_2\|S_3\|S_4$. The state is first initialized by a key and nonce, and then updated by associated data blocks and message blocks step by step. Hence, we use a superscript to denote the step number of the state as $S^t_0\|S^t_1\|S^t_2\|S^t_3\|S^t_4$, where $t$ is $-16$ before the initialization and $t=0$ after the initialization.

@@

The initialization of \cipher{MORUS1280} is slightly different from that of \cipher{MORUS640} due to the different register size and the two possible key size:
\begin{itemize}
  \item[-] When the key size is 128 bits, $K = K_{128} \|K_{128}$
  \item[-] When the key size is 256 bits, $K = K_{256}$
\end{itemize}
The state is initialized as
\begin{align*}
S^{-16}_0 &= \textrm{IV}_{128} \| 0^{128}, &
S^{-16}_1 &= K, &
S^{-16}_2 = 1^{256}, \\
S^{-16}_3 &= 0^{256}, &
S^{-16}_4 &= const_0\|const_1.
\end{align*}
After updating the state 16 times, the state is updated by $S^0_1 \leftarrow S^0_1 \oplus K$.

### 2018-05-09 morusAC_05_Full_trails.tex
propose introduction (poorly written)

\section{Trail for Full \cipher{MORUS}}

\subsection{Symmetrizing the Trail}
In the previous section, we have exhibited a linear trail on the reduced ciphers \cipher{MiniMORUS640} and \cipher{MiniMORUS1280}. We now turn to the full ciphers \cipher{MORUS640} and \cipher{MORUS1280}.
In order to build a trail on the full \cipher{MORUS}, we proceed exactly like we did for \cipher{MiniMORUS}, following the exact same path down to step and word rotation values, with one difference: in order to move from the one-word registers of \cipher{MiniMORUS} to the four-word registers of full \cipher{MORUS}, we make every term $S^t_{i,j}$ and $C^t_j$ symmetric, in the sense of \Cref{sec:introminimorus}.
That is, for every $S^t_{i,j}$ (resp. $C^t_j$) component in every trail fragment and every equation, we \emph{symmetrize} the term by adding in the terms $S^t_{i,j+w}$, $S^t_{i,j+2w}$, $S^t_{i,j+3w}$ (resp. $C^t_{j+w}$, $C^t_{j+2w}$, $C^t_{j+3w}$), where as usual $w$ denotes the word size. For example, if $w=32$ (for \cipher{MORUS640}), the term $S^3_{2,0}$ is symmetrized into:
\[
S^3_{2,0} \oplus S^3_{2,32} \oplus S^3_{2,64} \oplus S^3_{2,96}.
\]

@@

The symmetrized trail on full \cipher{MORUS} may be intuitively understood as consisting of four copies of the original trail on \cipher{MiniMORUS}. Indeed, the only difference between full \cipher{MORUS} (for either version of \cipher{MORUS}) and four independent copies of \cipher{MiniMORUS} comes from register-wise rotations, which permute words within a register. But as observed in \Cref{sec:introminimorus}, register-wise rotations leave symmetric linear combinations invariant; and so, insofar as we only ever use symmetric linear combinations on all registers along the trail, register-wise rotations have no effect.

>>>>

\section{Trail for Full \cipher{MORUS}}
\label{sec/fulltrails}

In the previous section, we have exhibited a linear trail on the reduced ciphers \cipher{MiniMORUS640} and \cipher{MiniMORUS1280}. We now turn to the full ciphers \cipher{MORUS640} and \cipher{MORUS1280}.

\subsection{Symmetrizing the Trail}
In order to build a trail on the full \cipher{MORUS}, we proceed exactly like we did for \cipher{MiniMORUS}, following the exact same path down to step and word rotation values, with one difference: in order to move from the one-word registers of \cipher{MiniMORUS} to the four-word registers of full \cipher{MORUS}, we make every term $S^t_{i,j}$ and $C^t_j$ symmetric, in the sense of \Cref{sec/introminimorus}.
That is, for every $S^t_{i,j}$ (resp. $C^t_j$) component in every trail fragment and every equation, we \emph{symmetrize} the term by adding in the terms $S^t_{i,j+w}$, $S^t_{i,j+2w}$, $S^t_{i,j+3w}$ (resp. $C^t_{j+w}$, $C^t_{j+2w}$, $C^t_{j+3w}$), where as usual $w$ denotes the word size. For example, if $w=32$ (for \cipher{MORUS640}), the term $S^3_{2,0}$ is symmetrized into:
\[
S^3_{2,0} \oplus S^3_{2,32} \oplus S^3_{2,64} \oplus S^3_{2,96}.
\]

@@

The symmetrized trail on full \cipher{MORUS} may be intuitively understood as consisting of four copies of the original trail on \cipher{MiniMORUS}. Indeed, the only difference between full \cipher{MORUS} (for either version of \cipher{MORUS}) and four independent copies of \cipher{MiniMORUS} comes from register-wise rotations, which permute words within a register. But as observed in \Cref{sec/introminimorus}, register-wise rotations leave symmetric linear combinations invariant; and so, insofar as we only ever use symmetric linear combinations on all registers along the trail, register-wise rotations have no effect.
